---
title: "nanonext - NNG Lightweight Messaging Library"
vignette: >
  %\VignetteIndexEntry{nanonext - NNG Lightweight Messaging Library}
  %\VignetteEngine{litedown::vignette}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```

### 1. Cross-language Exchange

`nanonext` provides a fast, reliable data interface between different programming languages where NNG has an implementation, including C, C++, Java, Python, Go, and Rust.

This messaging interface is lightweight, robust, and has limited points of failure. It enables:

- Communication between processes in the same or different languages
- Distributed computing across networks or on the same machine
- Real-time data pipelines where computation times exceed data frequency
- Modular software design following Unix philosophy

The following example demonstrates numerical data exchange between R and Python (NumPy).

Create socket in Python using the NNG binding 'pynng':

```{python}
#| label: rnpy1
import numpy as np
import pynng
socket = pynng.Pair0(listen="ipc:///tmp/nanonext.socket")
```

Create nano object in R using `nanonext`, then send a vector of 'doubles', specifying mode as 'raw':

```{r}
#| label: rnpy2
library(nanonext)
n <- nano("pair", dial = "ipc:///tmp/nanonext.socket")
n$send(c(1.1, 2.2, 3.3, 4.4, 5.5), mode = "raw")
```

Receive in Python as a NumPy array of 'floats', and send back to R:

```{python}
#| label: rnpy3
raw = socket.recv()
array = np.frombuffer(raw)
print(array)

msg = array.tobytes()
socket.send(msg)

socket.close()
```

Receive in R, specifying the receive mode as 'double':

```{r}
#| label: rnpy4
n$recv(mode = "double")

n$close()
```

### 2. Async and Concurrency

`nanonext` implements true async send and receive, leveraging NNG as a massively-scalable concurrency framework.

```{r}
#| label: async
s1 <- socket("pair", listen = "inproc://nano")
s2 <- socket("pair", dial = "inproc://nano")

```

`send_aio()` and `recv_aio()` return immediately with an 'Aio' object that performs operations asynchronously. Aio objects return an unresolved value while the operation is ongoing, then automatically resolve once complete.

```{r}
#| label: async4
# async receive requested, but no messages waiting yet
msg <- recv_aio(s2)
msg
msg$data

```

For 'sendAio' objects, the result is stored at `$result`. For 'recvAio' objects, the message is stored at `$data`.

```{r}
#| label: async2
res <- send_aio(s1, data.frame(a = 1, b = 2))
res
res$result

```
*Note: 0 indicates successful send - the message has been accepted by the socket for sending but may still be buffered within the system.*

```{r}
#| label: async3
# once a message is sent, the recvAio resolves automatically
msg$data

```

Use `unresolved()` in control flow to perform actions before or after Aio resolution without blocking.

```{r}
#| label: async5
msg <- recv_aio(s2)

# unresolved() checks resolution status
while (unresolved(msg)) {
  # perform other tasks while waiting
  send_aio(s1, "resolved")
  cat("unresolved")
}

# access resolved value
msg$data
```

Explicitly wait for completion with `call_aio()` (blocking).

```{r}
#| label: async7
# wait for completion and return resolved Aio
call_aio(msg)

# access resolved value (waiting if required):
call_aio(msg)$data

# or directly:
collect_aio(msg)

# or user-interruptible:
msg[]

close(s1)
close(s2)

```

### 3. Synchronisation Primitives

`nanonext` implements cross-platform synchronisation primitives from the NNG library, enabling synchronisation between NNG events and the main R execution thread.

Condition variables can signal events such as asynchronous receive completions and pipe events (connections established or dropped). Each condition variable has a value (counter) and flag (boolean). Signals increment the value; successful `wait()` or `until()` calls decrement it. A non-zero value allows waiting threads to continue.

This approach is more efficient than polling - consuming no resources while waiting and synchronising with zero latency.

**Example 1: Wait for connection**

```{r}
#| label: pipeevt
sock <- socket("pair", listen = "inproc://nanopipe")

cv <- cv()
cv_value(cv)

pipe_notify(sock, cv = cv, add = TRUE, remove = TRUE)

# wait(cv) would block until connection established

# for illustration:
sock2 <- socket("pair", dial = "inproc://nanopipe")

cv_value(cv) # incremented when pipe created

wait(cv) # does not block as cv value is non-zero

cv_value(cv) # decremented by wait()

close(sock2)

cv_value(cv) # incremented when pipe destroyed

close(sock)

```

**Example 2: Wait for message or disconnection**

```{r}
#| label: signalevt
sock <- socket("pair", listen = "inproc://nanosignal")
sock2 <- socket("pair", dial = "inproc://nanosignal")

cv <- cv()
cv_value(cv)

pipe_notify(sock, cv = cv, add = FALSE, remove = TRUE, flag = TRUE)

send(sock2, "this message will wake waiting thread")

r <- recv_aio(sock, cv = cv)

# wakes when async receive completes
wait(cv) || stop("peer disconnected")

r$data

close(sock)
close(sock2)

```

When `flag = TRUE` is set for pipe notifications, `wait()` returns FALSE for pipe events (rather than TRUE for message events). This distinguishes between disconnections and successful receives, something not possible using `call_aio()` alone.

This mechanism enables waiting simultaneously on multiple events while distinguishing between them. `pipe_notify()` can signal up to two condition variables per event for additional flexibility in concurrent applications.

### 4. TLS Secure Connections

Secure connections use NNG and Mbed TLS libraries. Enable them by:

1. Specifying a secure `tls+tcp://` or `wss://` URL
2. Passing a TLS configuration object to the 'tls' argument of `listen()` or `dial()`

Create TLS configurations with `tls_config()`:
- Client configuration: requires PEM-encoded CA certificate to verify server identity
- Server configuration: requires certificate and private key

Certificates may be supplied as files or character vectors. Valid X.509 certificates from Certificate Authorities are supported.

The convenience function `write_cert()` generates a 4096-bit RSA key pair and self-signed X.509 certificate. The 'cn' argument must match exactly the hostname/IP address of the URL (e.g., use '127.0.0.1' throughout, or 'localhost' throughout, not mixed).

```{r}
#| label: tlsconfig
cert <- write_cert(cn = "127.0.0.1")
str(cert)

ser <- tls_config(server = cert$server)
ser

cli <- tls_config(client = cert$client)
cli

s <- socket(listen = "tls+tcp://127.0.0.1:5558", tls = ser)
s1 <- socket(dial = "tls+tcp://127.0.0.1:5558", tls = cli)

# secure TLS connection established

close(s1)
close(s)

```

### 5. Request Reply Protocol

`nanonext` implements remote procedure calls (RPC) using NNG's req/rep protocol for distributed computing. Use this for computationally-expensive calculations or I/O-bound operations in separate server processes.

**[S] Server process:** `reply()` waits for a message, applies a function, and sends back the result. Started in a background 'mirai' process.

```{r}
#| label: rpcserver
m <- mirai::mirai({
  library(nanonext)
  rep <- socket("rep", listen = "tcp://127.0.0.1:6556")
  reply(context(rep), execute = rnorm, send_mode = "raw")
  Sys.sleep(2) # linger period to flush system socket send
})

```

**[C] Client process:** `request()` performs async send/receive, returning immediately with a `recvAio` object.

```{r}
#| label: rpcclient
library(nanonext)
req <- socket("req", dial = "tcp://127.0.0.1:6556")
aio <- request(context(req), data = 1e8, recv_mode = "double")

```
The client can now run additional code while the server processes the request.

```{r}
#| label: rpcclient2
# do more...
```

When the result is needed, call the recvAio using `call_aio()` to retrieve the value at `$data`.

```{r}
#| label: rpcclient3
call_aio(aio)$data |> str()
```

Since `call_aio()` blocks, alternatively query `aio$data` directly, which returns 'unresolved' (logical NA) if incomplete.

For server-side operations (e.g., writing to disk), calling or querying the value confirms completion and provides the function's return value (typically NULL or an exit code).

The [`mirai`](https://doi.org/10.5281/zenodo.7912722) package (<https://mirai.r-lib.org/>) uses `nanonext` as the back-end to provide asynchronous execution of arbitrary R code using the RPC model.

### 6. Publisher Subscriber Protocol

`nanonext` implements NNG's pub/sub protocol. Subscribers can subscribe to one or multiple topics broadcast by a publisher.

```{r}
#| label: pub
pub <- socket("pub", listen = "inproc://nanobroadcast")
sub <- socket("sub", dial = "inproc://nanobroadcast")

sub |> subscribe(topic = "examples")

pub |> send(c("examples", "this is an example"), mode = "raw")
sub |> recv(mode = "character")

pub |> send("examples at the start of a single text message", mode = "raw")
sub |> recv(mode = "character")

pub |> send(c("other", "this other topic will not be received"), mode = "raw")
sub |> recv(mode = "character")

# specify NULL to subscribe to ALL topics
sub |> subscribe(topic = NULL)
pub |> send(c("newTopic", "this is a new topic"), mode = "raw")
sub |> recv("character")

sub |> unsubscribe(topic = NULL)
pub |> send(c("newTopic", "this topic will now not be received"), mode = "raw")
sub |> recv("character")

# however the topics explicitly subscribed to are still received
pub |> send(c("examples will still be received"), mode = "raw")
sub |> recv(mode = "character")

```

The subscribed topic can be of any atomic type (not just character), allowing integer, double, logical, complex and raw vectors to be sent and received.

```{r}
#| label: pub2
sub |> subscribe(topic = 1)
pub |> send(c(1, 10, 10, 20), mode = "raw")
sub |> recv(mode = "double")
pub |> send(c(2, 10, 10, 20), mode = "raw")
sub |> recv(mode = "double")

close(pub)
close(sub)

```

### 7. Surveyor Respondent Protocol

Useful for service discovery and similar applications. A surveyor broadcasts a survey to all respondents, who may reply within a timeout period. Late responses are discarded.

```{r}
#| label: survey
sur <- socket("surveyor", listen = "inproc://nanoservice")
res1 <- socket("respondent", dial = "inproc://nanoservice")
res2 <- socket("respondent", dial = "inproc://nanoservice")

# sur sets a survey timeout, applying to this and subsequent surveys
sur |> survey_time(value = 500)

# sur sends a message and then requests 2 async receives
sur |> send("service check")
aio1 <- sur |> recv_aio()
aio2 <- sur |> recv_aio()

# res1 receives the message and replies using an aio send function
res1 |> recv()
res1 |> send_aio("res1")

# res2 receives the message but fails to reply
res2 |> recv()

# checking the aio - only the first will have resolved
aio1$data
aio2$data

# after the survey expires, the second resolves into a timeout error
msleep(500)
aio2$data

close(sur)
close(res1)
close(res2)

```

`msleep()` is an uninterruptible sleep function (using NNG) that takes a time in milliseconds.

The final value resolves to a timeout error (integer 5 classed as 'errorValue'). All error codes are classed as 'errorValue' for easy distinction from integer message values.

### 8. Options, Serialization and Statistics

Use `opt()` and `'opt<-'()` to get and set options on Sockets, Contexts, Streams, Listeners, or Dialers. See function documentation for available options.

To configure dialers or listeners after creation, specify `autostart = FALSE` (configuration cannot be changed after starting).

```{r}
#| label: opt
s <- socket(listen = "inproc://options", autostart = FALSE)

# no maximum message size
opt(s$listener[[1]], "recv-size-max")

# enforce maximum message size to protect against denial-of-service attacks
opt(s$listener[[1]], "recv-size-max") <- 8192L

opt(s$listener[[1]], "recv-size-max")

start(s$listener[[1]])

```

The special write-only option 'serial' sets a serialization configuration via `serial_config()`. This registers custom functions for serializing/unserializing reference objects using R's 'refhook' system, enabling transparent send/receive with mode 'serial'. Configurations apply to the Socket and all Contexts created from it.

```{r}
#| label: serial
serial <- serial_config("obj_class", function(x) serialize(x, NULL), unserialize)
opt(s, "serial") <- serial

close(s)
```

Use `stat()` to access NNG's statistics framework. Query Sockets, Listeners, or Dialers for statistics such as connection attempts and current connections. See function documentation for available statistics.

```{r}
#| label: stat
s <- socket(listen = "inproc://stat")

# no active connections (pipes)
stat(s, "pipes")

s1 <- socket(dial = "inproc://stat")

# one now that the dialer has connected
stat(s, "pipes")

close(s)
```
